{\bf Intrusion Detection through Provenance-Based Histograms}

### 2013-11-24 writeup/paper.tex
Abstract and Related Works

\begin{abstract}
TODO
\end{abstract}

@@

Given the extensive amount of data PASS collects on file-file, file-process, and process-process dependencies, it seems plausible to detect intrusions based on provenance data collected by PASS. Tariq et. al. generalize Somayaji's sequences of system calls to sequences of provenance system events \cite{correlated-anomalies} to develop an intrusion detection system (IDS) in a distributed system. Their system uses bloom filters to store hashes of $k$-tuples--representing a sequence of provenance events--and then use these bit vectors to correlate anomalies across multiple hosts. King et. al. have developed Backtrack \cite{backtrack}, a modified Linux kernel that tracks dependencies between operating system objects (files, processes, file names) in a similar fashion to PASS. Given an intrusion, Backtrack builds a backwards casual graph to determine the entry point of an intrusion, and they have extended Backtrack with forward causal graphs to determine possibly tainted files, processes, and hosts from an intrusion in a distributed system \cite{multihost}. However, their system does not find a way of detecting an intrusion. Lei et. al. build similar process-file dependency trees, which they call access trees, and define a compatibility function to compare access trees with the goal of predicting future file accesses based on the current access tree pattern \cite{fileprefetch}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN
%

@@

\subsection{Evaluation Methodology}
In order to test our techniques, we needed to run “exploits” on our PASS-enabled machine. We looked to prior work to get a sense of the different types of techniques commonly used to exploit security vulnerabilities. Commonly, remote network techniques [CITE DUKE PAPER] are used to broach insecure systems. Metasploit [CITE] is a software package designed to facilitate running exploits on insecure system. Searching the Metasploit database, as well as the online Exploit Database (exploit-db.com) [CITE], we found that most successful techniques involve exploiting vulnerabilities in software that a target system is running, as opposed to vulnerabilities in the target system itself. Put another way, we sought to exploit software packages, like mcrypt and gcc, not the Linux kernel itself.

We found that many of the popular exploits involved buffer overflows and arbitrary code execution. Programming bugs in a piece of software allowed a malicious user, sending well-crafted data, to run arbitrary code on the target system. In particular, the mcrypt encryption package had a long standing buffer overflow, which would allow a malicious user to run their own code when mcrypt tried to decrypt the crafted exploit file. Due to limitations in our Virtual Machine set up, we were unable to exploit the vulnerable version of mcrypt, much like [CITE DUKE]. To emulate this exploit, we designed our own program, which would decrypt (via ROT13) an input file. When sent a particular “exploit file” which contained malicious lines, our program would stop decrypting and start running the code specified in the file.

@@

\bibitem{backtrack}
\textsc{King, S. T. and Chen, P. M.} Backtracking Intrusions. In {\em SOSP'03 Proceedings of the nineteenth ACM symposium on Operating systems principles} (December 2003).

>>>>

\begin{abstract}
Data-driven intrusion detection is a common system security technique. Here we present a method for generating a statistical model based approach to intrusion detection, using provenance data generated by a provenance-aware modified file system. Our analyses reveal that provenance data can be directly used to detect intrusions, and an in-depth analysis reveals which data is most useful in distinguishing between safe and malicious programs and program executions. Our histogram based approach requires no prior intuition about the types of exploits potentially being run on a system, and identifies suspicious behavior based on deviations from prior normal system behavior.
\end{abstract}

@@

Given the extensive amount of data PASS collects on file-file, file-process, and process-process dependencies, it seems plausible to detect intrusions based on provenance data collected by PASS. Tariq et. al. generalize Somayaji's sequences of system calls to sequences of provenance system events \cite{correlated-anomalies} to develop an intrusion detection system (IDS) in a distributed system. Their system uses bloom filters to store hashes of $k$-tuples--representing a sequence of provenance events--and then use these bit vectors to correlate anomalies across multiple hosts. King et. al. have developed Backtracker \cite{backtracker}, a modified Linux kernel that tracks dependencies between operating system objects (files, processes, file names) in a similar fashion to PASS. Given an intrusion, Backtracker builds a backwards casual graph to determine the entry point of an intrusion, and they have extended Backtracker with forward causal graphs to determine possibly tainted files, processes, and hosts from an intrusion in a distributed system \cite{multihost}. However, their system does not find a way of detecting an intrusion. Lei et. al. build similar process-file dependency trees, which they call access trees, and define a compatibility function to compare access trees with the goal of predicting future file accesses based on the current access tree pattern \cite{fileprefetch}.

@@

\section{Background and Related Works}
\subsection{PASS}
Provenance data is meta-data detailing the creation, transformation, and flow of data. PASS \cite{pass, passv2} is a provenance-aware storage system, designed to generate and track this provenance data seamlessly. The PASS system automatically creates and updates provenance data fore every file that was created or modified the volume. The PASS provenance data is very rich data. For example, the provenance for a data file that sent in a zip file would reveal both the name of the program that unzipped the file, as well as the web browser through which the zip file was downloaded! It is our belief that this level of specificity will lend itself to detecting patterns in program usage.

The PASS system runs with minimal time overhead, and a reasonable space overhead (XXX cite real numbers).
\subsection{Intrusion Detection}
Intrusion detection in live systems is a common problem in the systems community. There are two primary components to an intrusion detection system (IDS): (1) the data used to fuel the system's analysis, and (2) the process in which this data is used.

Somayaji et al. \cite{somayaji} aimed to detect intrusions by keeping track of all systems calls that occur during a process's execution time. After collecting this data over a long period of "normal" time, if a proces exhibits system call patterns that deviate from the "normal" patterns for the process, their system flags it as potentialy dangerous, and slows its execution. In this way, a program that continually deviates from normal behavior will be descheduled entirely.

Tariq et al. \cite{correlated-anomalies} extend the work of Somayaji and others, designing a system to run an IDS on a set of distributed nodes. They use the IDS of other researchers, distilling a programs execution into a set of values (a "k-tuple") used to identify information of interest regarding the program's exeuction. Our system is designed to run offline, so we run our algorithms on the full set of provenance data. While this approach is slower, it allows us to take in the full picture of what occurred during a process's execution, without having to preselect the information we think will be useful. This approach is especially helpful because different information leads to the detection of different types of intrusions.

Backtracker \cite{backtracker} is a system most similar to ours in design. It, like PASS, collects data while the system is running, and like our system analyzes it offline for potential intrusions. Whereas PASS collects data directly at the operating system level (using a modified file system), Backtracker runs its target OS inside of a virtual machine, and uses a modified virtual machine monitor to collect data. The data they collect is very similar to the data PASS generates, and their exploits and analysis techniques directly inspired our work. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN
%

@@

\subsection{Evaluation Methodology}
In order to test our techniques, we needed to run “exploits” on our PASS-enabled machine. We looked to prior work to get a sense of the different types of techniques commonly used to exploit security vulnerabilities. Commonly, remote network techniques (as seen in Backtracker \cite{backtracker}) are used to broach insecure systems. Metasploit [CITE] is a software package designed to facilitate running exploits on insecure system. Searching the Metasploit database, as well as the online Exploit Database (exploit-db.com) [CITE], we found that most successful techniques involve exploiting vulnerabilities in software that a target system is running, as opposed to vulnerabilities in the target system itself. Put another way, we sought to exploit software packages, like mcrypt and gcc, not the Linux kernel itself.

We found that many of the popular exploits involved buffer overflows and arbitrary code execution. Programming bugs in a piece of software allowed a malicious user, sending well-crafted data, to run arbitrary code on the target system. In particular, the mcrypt encryption package had a long standing buffer overflow, which would allow a malicious user to run their own code when mcrypt tried to decrypt the crafted exploit file. Due to limitations in our Virtual Machine set up, we were unable to exploit the vulnerable version of mcrypt, much like the authors of Backtracker \cite{backtracker} found. To emulate this exploit, we designed our own program, which would decrypt (via ROT13) an input file. When sent a particular “exploit file” which contained malicious lines, our program would stop decrypting and start running the code specified in the file.

@@

\bibitem{backtracker}
\textsc{King, S. T. and Chen, P. M.} Backtracking Intrusions. In {\em SOSP'03 Proceedings of the nineteenth ACM symposium on Operating systems principles} (December 2003).
